\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/experiments.tex
Move commands so they can be used in any sec
paper: clean up complexity section
Add performance and smape figures to text

\newcommand{\demorows}{10175}
\newcommand{\labexrows}{9813}
\newcommand{\acsbaseresultminutes}{355 minutes}
\newcommand{\acsbaseresulthours}{just under 6 hours}
\newcommand{\acsimputedbzeroresult}{4 seconds}
\newcommand{\acsimputedboneresult}{1 second}

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running
imputation on the base tables and executing the query on this cleaned base data \todo{SMAPE citation}.
Each query was separately run 200 times in both settings and then results from the two approaches were paired up and compared tuple-wise.
We average tuple-wise absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
We can see that optimizing for quality indeed reduces the SMAPE of query results.
In general, the SMAPE relative to the base-imputation approach are low in all cases\todo{give actual value}, indicating that on-the-fly imputation produces
similar results to imputation at the base tables.

>>>>

@@

\Cref{table:smape} shows the Symmetric-Mean-Absolute-Percentage-Error (SMAPE) for \ProjectName{}'s query results when compared to running
imputation on the base tables and executing the query on this cleaned base data \todo{SMAPE citation}.
Each query was separately run 200 times in both settings and then results from the two approaches were paired up and compared tuple-wise.
We average tuple-wise absolute percentage deviations within each iteration of a query, and we report this value averaged over all iterations.
We can see that optimizing for quality indeed reduces the SMAPE of query results.
In general, the SMAPE relative to the base-imputation approach are low in all cases --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent ---, indicating that on-the-fly imputation produces
similar results to imputation at the base tables.

### 2017-02-28 paper/introduction.tex
Add performance and smape figures to text

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression algorithm takes about XXX minutes to train and run
on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes to impute missing values on a 1\% sample of an XXX row database.
\srm{Make numbers above consistent with experiments.}
A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
queries that execute in a fraction of the time\todo{add actual performance figures here} it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the complete imputation approach, as indicated by low\todo{add actual value here} Symmetric-Mean-Absolute-Percentage-Errors in our empirical results
over real-world data sets (see~\Cref{sec:experiments} for details).

>>>>

% Solution
One common approach is to manually replace missing values using some sort of statistical or predictive model based on other values in the table and record.
This process is called \emph{imputation}.
In this paper, we present a new system, called \ProjectName{}, which is the designed to selectively apply imputation to a subset of records, on-the-fly, during query execution.  \emph{The key insight behind \ProjectName{} is that imputation
only needs to be performed on the data relevant to a particular query and
that this subset is generally much smaller than the entire database.}  
% Background
While traditional imputation methods work over the entire data set and replace all missing values, running a sophisticated imputation algorithm over a large data set can be very expensive---our experiments show that even a relatively simple regression tree algorithm takes \acsbaseresulthours{} to train and run
on a 600K row database; prior work~\cite{akande2015empirical} reports times of 45 minutes for multiple-imputation of missing values on a 1\% sample of the same database on a desktop computer.\todo{Explain the discrepancy between 45 * 100 and 355? They use L=10 datasets and unknown hardware. Or drop this comparison entirely?}

A simpler approach might drop all rows with missing values, but not only can this introduce bias into the results, but this may result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement  imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in placement of imputation operations
in an optimal way into query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue standard SQL queries over a data set, even if that data has missing values, and get
queries that execute between \lowxalphazero{} and \highxalphaoneexacs{} times as fast as it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the complete imputation approach, as indicated by low Symmetric-Mean-Absolute-Percentage-Errors --- between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent --- in our empirical results
over real-world data sets (see~\Cref{sec:experiments} for details).

### 2017-02-28 paper/main.tex
pass on intro
paper: pass over experiments
move related work to end
comments
Move commands so they can be used in any sec
Add performance and smape figures to text

\documentclass{vldb}
\usepackage{balance}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage[subtle]{savetrees}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=numeric,firstinits=true]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx} % includegraphics
\usepackage{listings} % code samples with syntax hl
\usepackage{stmaryrd}
\usepackage{qtree} % for simple tree diagrams
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{hyperref} % load after other packages
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{cleveref}

\addbibresource{references.bib}

@@

% listings settings
\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
  basicstyle=\ttfamily\footnotesize,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{red},    % comment style
  %frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
%  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
%  numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
%  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mauve},     % string literal style
  tabsize=2,	                   % sets default tabsize to 2 spaces
  language=SQL
}

@@

\newcommand{\ProjectName}{ImputeDB}
\newcommand{\doms}[2]{#1 \succ #2}

@@

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} #1\\#2}}}}
\begin{document}

@@

\begin{abstract}
Missing values abound in data analysis and often present a serious challenge for usability of data. Users
are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
by  applying a potentially-expensive imputation strategy. We show that it is possible
to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
for each query. This allows users to immediately explore
their data, while the system picks the optimal placement of imputation operations. This placement is based
on the specific query issued and a user-provided parameter that trades off data
quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
generates query plans that execute between \todo{add [low x - high x] range} as fast as imputing at the base
tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \todo{add SMAPE range}  
when compared to standard full-table imputation approach.
\end{abstract}

\input{introduction}
\input{examples}
\input{relatedwork}
\input{algorithm}
\input{experiments}
\input{conclusion}
\balance
\printbibliography

>>>>

\documentclass{vldb}
\usepackage{balance}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage[subtle]{savetrees}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=numeric,firstinits=true]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx} % includegraphics
\usepackage{listings} % code samples with syntax hl
\usepackage{stmaryrd}
\usepackage{qtree} % for simple tree diagrams
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{hyperref} % load after other packages
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{cleveref}
\usepackage{xspace}

\newcommand{\srm}[1]{\textcolor{red}{SAM:#1}}
\addbibresource{references.bib}

@@

% listings settings
\lstset{ %
  backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}; should come as last argument
  basicstyle=\ttfamily\footnotesize,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{red},    % comment style
  %frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
%  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
%  numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
%  stepnumber=2,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{red},     % string literal style
  tabsize=2,	                   % sets default tabsize to 2 spaces
  language=SQL
}

@@

\newcommand{\ProjectName}{ImputeDB\xspace}
\newcommand{\sys}{\ProjectName}
\newcommand{\doms}[2]{#1 \succ #2}

@@

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} #1\\#2}}}}

% results that can be used in any section
\newcommand{\demorows}{10175}
\newcommand{\labexrows}{9813}
\newcommand{\acsbaseresultminutes}{355 minutes}
\newcommand{\acsbaseresulthours}{just under 6 hours}
\newcommand{\acsimputedbzeroresult}{4 seconds}
\newcommand{\acsimputedboneresult}{1 second}
\newcommand{\lowxalphazero}{10}
\newcommand{\lowsmapealphazero}{0}
\newcommand{\highxalphaoneexacs}{1000}
\newcommand{\highsmapealphaoneexacs}{24}
\newcommand{\highxalphaone}{16000}

\begin{document}

@@

\begin{abstract}
Missing values abound in data analysis and often present a serious challenge for usability of data. Users
are forced to pick between removing tuples with missing values or creating a clean duplicate of their data
by  applying a potentially-expensive imputation strategy. We show that it is possible
to incorporate imputation into a cost-based query optimizer, performing imputation on the fly
for each query. This allows users to immediately explore
their data, while the system picks the optimal placement of imputation operations. This placement is based
on the specific query issued and a user-provided parameter that trades off data
quality and runtime performance. We incorporate the imputation operators into the relational algebra, allowing
for traditional optimizations to take place. Our experiments evaluate this approach on two real-world survey-based datasets
from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}. Our results show that the optimizer
generates query plans that execute between \lowxalphazero{} and \highxalphaoneexacs{} times as fast as imputing at the base
tables and then executing queries, making exploratory data analysis using our system feasible. Furthermore, we show that the query results
from on-the-fly imputation have Symmetric-Mean-Absolute-Percentage-Error between \lowsmapealphazero{} and \highsmapealphaoneexacs{} percent
when compared to standard full-table imputation approach.
\end{abstract}

\input{introduction}
\input{examples}
\input{algorithm}
\input{experiments}
\input{relatedwork}
\input{conclusion}
\balance
\bibstyle{abbrv}
\printbibliography
